---
title: "Group Task"
author: "Nzambuli Daniel", "Chesia Anika", "Brandon Gem", "Waimiri Romeo"
date: "2024-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Libraries to Model SEIR-D with Children and Adults

```{r}
library(deSolve)
library(tidyverse)
library(DiagrammeR)
```

## States

```{r}
grViz("digraph flowchart {
      graph [layout = dot,
       rankdir = LR]
      node [fontname = Helvetica, shape = rectangle,]
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3'] 
      tab4 [label = '@@4']
      tab5 [label = '@@5']
      tab6 [label = '@@6']
      tab7 [label = '@@7']
      tab8 [label = '@@8']
      tab9 [label = '@@9']
      # edge definitions with the node IDs
      tab1 -> tab2 [label = 'beta child'];
      tab1 -> tab3 [label = 'beta adult'];
      tab2 -> tab4 [label = 'kapa child'];
      tab4 -> tab6 [label = 'mu child'];
      tab4 -> tab8 [label = 'gamma child'];
      tab3 -> tab5 [label = 'kapa adult'];
      tab5 -> tab7 [label = 'mu adult'];
      tab5 -> tab9 [label = 'gamma adult'];
      }

      [1]: 'Susceptible'
      [2]: 'Exposed Children'
      [3]: 'Exposed Adults'
      [4]: 'Infected Children'
      [5]: 'Infected Adult'
      [6]: 'Dead Children'
      [7]: 'Dead Adults'
      [8]: 'Recovered Children'
      [9]: 'Recovered Adults'
      ")
```

The SEIRD model extends the SEIR model by adding a compartment for individuals who have succumbed to the disease. The basic flow of individuals between compartments in the SEIRD model is as follows:

1.  S (Susceptible): Individuals who are not infected but can become exposed to the virus.
2.  E (Exposed): Individuals who have been exposed to the virus but are not yet infectious.
3.  I (Infectious): Individuals who are currently infected and can spread the disease to others.
4.  R (Recovered): Individuals who have recovered from the infection and are assumed to be immune.
5.  D (Dead): Individuals who have died as a result of the disease.

## The Differential Equations

$$
\frac{dS(t)}{dt} = \frac{-\beta S(t)I(t)}{N_{pop}}\\ \\
\frac{dE(t)}{dt} = \frac{+\beta S(t)I(t)}{N_{pop}} - \kappa E(t)\\ \\
\frac{dI(t)}{dt} = -\gamma I(t) + \kappa E(t)\\ \\
\frac{dR(t)}{dt} = +\gamma(1-\mu_h) min(\alpha I(t), I_h) + \gamma(1 - \mu_n)max(0, \alpha I(t), I_h) + \gamma(1-\alpha) I\\ \\
\frac{dD(t)}{dt} = +\gamma \mu_h min(\alpha I(t), I_h) + \gamma \mu_n max(0, \alpha I(t), I_h)\\ \\
\frac{dN_{pop}}{dt} = \frac{dS(t)}{dt} + \frac{dE(t)}{dt} + \frac{dI(t)}{dt} + \frac{dR(t)}{dt} + \frac{dD(t)}{dt} = 0
$$

## Initial Set-up
