---
title: "SEIR with Intervention Assignment"
author: "Nzambuli Daniel"
date: "2024-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task

## Information

**Parameters of Interest:** : Contact Rate (7), Transmission Probability (0.07), infectiousness Period (8), latency Period (30)

**Disease Dynamic Parameters and Initial Values**: Exposed (5000), Susceptible (9999), Infected (1), Recovered (0)

**State Values**

-   Susceptible: susceptible/Total population

-   Exposed: exposed/Total population

-   Infected: infected/Total population

-   Recovered: recovered/Total population

-   Time Period for Monitoring: 50 days

## Goal

1.  Present the Schematic Framework to Conceptualize Compartments (Sub-populations) when Modeling Age-Groups:
    -   **Objective:** Model age-groups, considering that susceptible children experience the force of infection, while susceptible adults become infected at a constant rate.
    -   **Assumptions:**
        1.  Children make 13 contacts per day on average, with 7 contacts being with other children.
        2.  Adults make 11 contacts per day on average, with 10 contacts being with other adults.
    -   **Schematic Framework:** Illustrate the compartments and interactions between age-groups.
2.  Simulate the SEIR Epidemic and Plot Different Sub-populations:
    -   **Objective:** Implement the SEIR model based on the provided parameters and initial values.
    -   **Steps:**
        1.  Define the SEIR model equations incorporating age-group-specific contact rates.
        2.  Utilize the provided initial values to simulate the epidemic over a 50-day period.
        3.  Plot the trajectories of Susceptible, Exposed, Infected, and Recovered sub-populations over time.
3.  Evaluate and Discuss Changes with Vaccination:
    -   **Objective:** Assume a vaccine with 100% effectiveness against acquiring infections and analyze its impact on sub-populations.
    -   **Steps:**
        1.  Introduce vaccination into the model, varying vaccination percentages (10%, 20%, 30%, 50%).
        2.  Simulate and observe changes in the trajectories of Susceptible, Exposed, Infected, and Recovered sub-populations.
        3.  Discuss how different vaccination rates influence the epidemic dynamics

::: {style="background-color: #f6bebo;"}
## Reminder

-   **Susceptible** – individuals who have not been exposed to the virus

-   **Exposed** – individuals exposed to the virus, but not yet infectious

-   **Infectious** – exposed individuals who go on to become infected

-   **Recovered** – infectious individuals who recover and become immune to the virus

-   **infectiousness period** – `time interval during which a host (individual or patient) is infectious`, i.e. capable of directly or indirectly transmitting pathogenic infectious agents or pathogens to another susceptible host.

**Population** $N$ – sum of the individuals in the four compartments.

1.  $\beta\ -\ Beta$

**Transmission coefficient** – The average number of `infectious contacts` an infected infectious individual makes each time period. *High beta* means that the disease has `more opportunity` to spread.

$$ contact\ rate * transmission\ probability $$

2.  $\sigma\ -\ Sigma$ or $\delta\ Delta$

**Exposed individuals become infectious** – The reciprocal of the average time it takes to become infectious. Exposion rate.

$$\frac{1}{latency\ period}$$

3.  $\gamma\ -\ Gamma$

**Rate of Recovery** – the reciprocal of the average time it takes to recover.

$$ \frac{1}{infectious\ period} $$

4\. $\mu\ -\ miu$

**Mortality rate** – mortality rate of infectious individuals.

the higher the $\mu$ the more deadly the disease is

**Reproduction Number**

$$ Recovery\ Number\ (RO) = \frac{Beta}{Gamma(recovery\ rate)}=\frac{\beta}{\gamma} $$
:::

# Working

```{r}
library(deSolve)
library(tidyverse)
library(DiagrammeR)
```

### States

```{r}
grViz("digraph flowchart {
      graph [layout = dot,
       rankdir = LR]
      node [fontname = Helvetica, shape = rectangle,]
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3'] 
      tab4 [label = '@@4']
      # edge definitions with the node IDs
      tab1 -> tab2;
      tab2 -> tab3;
      tab3 -> tab4;
      }

      [1]: 'Susceptible'
      [2]: 'Exposed'
      [3]: 'Infected'
      [4]: 'Recovered'
      ")
```

## Parameters

> Disease parameters
>
> -   Contact Rate (7)
>
> -   Transmission Probability (0.07)
>
> -   Infectiousness Period (8)
>
> -   Latency Period (30)
>
> Population parameters
>
> -   Exposed (5000)
>
> -   Susceptible (9999)
>
> -   Infected (1)
>
> -   Recovered (0)

::: {style="background-color: #d3d4d4;"}
1.  Children make 13 contacts per day on average, with 7 contacts being with other children.
2.  Adults make 11 contacts per day on average, with 10 contacts being with other adults.
3.  Children are 30% of the population.
:::

discoveries$$
N = N_{adult} + N_{children}\\
N_{children} = 0.3 N
$$

-   vaccination `does not` modify **exposure** and **infection**

-   susceptible children are expressed as $S_x$

$$
s_x = propotion\ *\ succeptible
$$

-   introducing intervention $v$ to $i$ % of the population creates a vaccinated population $v = i\ *\ N$

-   susceptible children $s_c$ needs to be modified to become

$$
S_c = 0.3 * 9999\\
S_c = S_c - (0.3 * V_c)
$$

```{r}
# initial population data
Exposed = 5000
Susceptible = 9999
Infected = 1
Recovered = 0

N = Exposed + Susceptible + Infected + Recovered
```

### Disease Parameters

```{r}
Contact.Rate = 7
Transmission.Probability = 0.07
Infectiousness.Period = 8
Latency.Period = 30

# gamma beta delta
gamma = 1/Infectiousness.Period
beta = Contact.Rate * Transmission.Probability
delta = 1/Latency.Period
```

### Interventions and Division of Population

```{r}
inter.1 = 0.1
inter.2 = 0.2
inter.3 = 0.3
inter.5 = 0.5

# percentage child  and adult
child.perc = 0.3
adult.perc = 0.7

# dividing into adult and children

# children
s.child = child.perc * Susceptible
e.child = child.perc * Exposed
i.child = child.perc * Infected
r.child = 0

# adults
s.adult = adult.perc * Susceptible
e.adult = adult.perc * Exposed
i.adult = adult.perc * Infected
r.adult = 0

```

#### Introducing Vaccination

**Remember** vaccination doesn't change `exposure` and `infection`

```{r}
# introducing vaccination
vac.1 = inter.1 * N

# child intervention introduced
vac.child = child.perc * vac.1
s.V.child = s.child - vac.child
e.v.child = e.child
i.v.child = i.child
r.v.child = r.child + vac.child

# adult intervention introduced
vac.adult = adult.perc * vac.1
s.v.adult = s.adult - vac.adult
e.v.adult = e.adult
i.v.adult = i.adult
r.v.adult = r.adult + vac.adult
```

## SEIR function

#### Beta for individual interaction

```{r}
c2c.beta = (7 * Transmission.Probability)/ Infectiousness.Period
c2a.beta = (6 * Transmission.Probability)/ Infectiousness.Period

# adult 
a2a.beta = (10 * Transmission.Probability)/ Infectiousness.Period
a2c.beta = (1 * Transmission.Probability)/ Infectiousness.Period
```

```{r}
initial_state = c(s.V.child, s.v.adult,
                  e.v.child, e.v.adult,
                  i.v.child, i.v.adult,
                  r.v.child, r.v.adult)
parameters = c(gamma = gamma,
               c2c.beta, c2a.beta, a2a.beta, a2c.beta,
               delta = delta)
time = seq(0, 100, 1)
```

### The function

```{r}
seir.func = function(time, state, parameters){
  S1 = state[1]
  S2 = state[2]
  E1 = state[3]
  E2 = state[4]
  I1 = state[5]
  I2 = state[6]
  R1 = state[7]
  R2 = state[8]
  with(as.list(parameters),{
    # number of children
    nc = S1 + E1 + I1 + R1
    # number of adults
    na = S2 + E2 + I2 + R2
    
    # derivative functions
    dS1 = -(c2c.beta + c2a.beta) * (S1 * I1)/nc
    dE1 = (c2c.beta + c2a.beta) * ((S1 * I1)/nc) - delta * E1
    dI1 = delta * E1 - gamma *I1
    dR1 = gamma * I1
    
    # adults
    dS2 = -(a2a.beta + a2c.beta) * (S2 * I2)/na
    dE2 = (a2a.beta + a2c.beta) * ((S2 * I2)/na) - delta * E2
    dI2 = delta * E2 - gamma *I2
    dR2 = gamma * I2
    
    result = c(dS1, dS2,
               dE1, dE2,
               dI1, dI2,
               dR1, dR2)
    return(list(result))
  })

  }
```

## Generating Output

```{r}
seir_10_out = as.data.frame(lsoda(initial_state,time,seir.func,parameters))
seir_10_out
```

```{r}
colnames(seir_10_out) = c("time","susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult")
```

### converting to long format

```{r}
seir_10_long = seir_10_out %>% pivot_longer(cols = c("susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult"),names_to = "state",values_to = "value"
)
seir_10_long
```

### Plot for 10% intervention

```{r}
ggplot(data = seir_10_long,aes(x = time, y = value, colour = state, group = state)) + geom_line() + labs(colour = "Compartment",title = "SEIR at 10% intervention",x="Time (days)",y = "Number of people")+theme_minimal()
```

> At `10 %` intervention there is a near intersection between infected adults and recovered adults.
>
> A hypothesis of increasing vaccination may lead to an equilibrium developing is developed and testing done.

# 20 %

#### Introducing Vaccination

**Remember** vaccination doesn't change `exposure` and `infection`

```{r}
# introducing vaccination 
vac.2 = inter.2 * N  
# child intervention introduced 
vac.child = child.perc * vac.2 
s.V.child = s.child - vac.child 
e.v.child = e.child 
i.v.child = i.child 
r.v.child = r.child + vac.child  
# adult intervention introduced 
vac.adult = adult.perc * vac.2
s.v.adult = s.adult - vac.adult 
e.v.adult = e.adult 
i.v.adult = i.adult 
r.v.adult = r.adult + vac.adult
```

## Generating Output

```{r}
seir_20_out = as.data.frame(lsoda(initial_state,time,seir.func,parameters)) 
seir_20_out
```

```{r}
colnames(seir_20_out) = c("time","susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult")
```

### converting to long format

```{r}
seir_20_long = seir_20_out %>% pivot_longer(cols = c("susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult"),names_to = "state",values_to = "value" ) 
seir_20_long
```

### Plot for 20% intervention

```{r}
ggplot(data = seir_20_long,aes(x = time, y = value, colour = state, group = state)) + geom_line() + labs(colour = "Compartment",title = "SEIR at 20% intervention",x="Time (days)",y = "Number of people")+theme_minimal()
```

> At `20 %` intervention infected children and adults never reach `1000` individuals and their number reduces from `day 12`
>
> Adult recovery has a steeper curve than child recovery from `day 12` as illustrated in the low infection rates.
>
> The exposed adults reduce in an almost `log` rate

# 30 %

#### Introducing Vaccination

**Remember** vaccination doesn't change `exposure` and `infection`

```{r}
# introducing vaccination  
vac.3 = inter.3 * N   
# child intervention introduced  
vac.child = child.perc * vac.3  
s.V.child = s.child - vac.child  
e.v.child = e.child  
i.v.child = i.child  
r.v.child = r.child + vac.child   
# adult intervention introduced  
vac.adult = adult.perc * vac.3 
s.v.adult = s.adult - vac.adult  
e.v.adult = e.adult  
i.v.adult = i.adult  
r.v.adult = r.adult + vac.adult
```

## Generating Output

```{r}
seir_30_out = as.data.frame(lsoda(initial_state,time,seir.func,parameters))  
seir_30_out
```

```{r}
colnames(seir_30_out) = c("time","susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult")
```

### converting to long format

```{r}
seir_30_long = seir_30_out %>% pivot_longer(cols = c("susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult"),names_to = "state",values_to = "value" )  
seir_30_long
```

### Plot for 30% intervention

```{r}
ggplot(data = seir_30_long,aes(x = time, y = value, colour = state, group = state)) + geom_line() + labs(colour = "Compartment",title = "SEIR at 30% intervention",x="Time (days)",y = "Number of people")+theme_minimal()
```

> Increasing the vaccination by `10` % seems to make the graphs `less steep` this is taken to indicate the effects of the epidemic are reducing with more vaccination

# 50 %

#### Introducing Vaccination

**Remember** vaccination doesn't change `exposure` and `infection`

```{r}
# introducing vaccination  
vac.5 = inter.5 * N   
# child intervention introduced  
vac.child = child.perc * vac.5  
s.V.child = s.child - vac.child  
e.v.child = e.child  
i.v.child = i.child  
r.v.child = r.child + vac.child   
# adult intervention introduced  
vac.adult = adult.perc * vac.5 
s.v.adult = s.adult - vac.adult  
e.v.adult = e.adult  
i.v.adult = i.adult  
r.v.adult = r.adult + vac.adult
```

## Generating Output

```{r}
seir_50_out = as.data.frame(lsoda(initial_state,time,seir.func,parameters))  
seir_50_out
```

```{r}
colnames(seir_50_out) = c("time","susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult")
```

### converting to long format

```{r}
seir_50_long = seir_50_out %>% pivot_longer(cols = c("susc.child", "susc.adult","exposed.child", "exposed.adult", "infected.child", "infected.adult","recovered.child", "recovered.adult"),names_to = "state",values_to = "value" )  
seir_50_long
```

### Plot for 50% intervention

```{r}
ggplot(data = seir_50_long,aes(x = time, y = value, colour = state, group = state)) + geom_line() + labs(colour = "Compartment",title = "SEIR at 50% intervention",x="Time (days)",y = "Number of people")+theme_minimal()
```

> There is minimal change in this case.
